---
title: "writeup"
output: github_document
---
# 01-roulette
### Zhuoyi Zhan
### September 06, 2021



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(dev = 'svg') # set output device to svg
```

## Introduction

In this question I want to answer the question to decide if Martingale strategy is always profitable. The purpose is to use programming to simulate real-life roulette, calculate average earning of the Martingale betting strategy and see if different input parameters would impact the average earnings. 

## 1.1 Backgound

Roulette is a popular gambling game. Starting at $1, if player guess right, he/she will get her money back and win double the bet. The Martingale is a common betting strategy that seems to guarantee return. When player lose,he/she doubles the amount of bet; if win, restart with 1. When stopping rule is met, player stops. We set starting budget, winning threshold, maximum wager, and maximum game to calculate average earning by repeating the simulations.

| Parameter         | value | 
|:-----------------:|:-----:|
| starting budget   |  $200 |
| winning threshold |  $300 |
| maximum wager     |  $100 |
| maximum game      |  1000 |


## 2 Methods

In this simulation, functions help to accomplish tasks like spinning wheels, recording outcomes and calculating budget.

```{r}
library(dplyr)
library(tidyverse)
 function(x) x[length(x)] 
```

```{r}
#This step we are trying to  simulate spinning the roulette. sample() is a function can take a random sample from the data object. In this case, yield a result from 18 red, 18 black, and 2 green
single_spin <- function(){
  possible_outcomes <- c(rep("red",18), rep("black",18), rep("green",2))
  sample(possible_outcomes, 1)
}

#This step is trying to simulate the betting strategy. It receives input of previous wager, last spinning result, maximum wager that casino set and how much much palyer left with. If last spin player get red, then keep bet $1 on red, else, 2 times the previous wager but it should not exceed player's current budget or maximum wager allowed.

martingale_wager <- function(
  previous_wager
  , previous_outcome
  , max_wager
  , current_budget
){
  if(previous_outcome == "red") return(1)
  min(2*previous_wager, max_wager, current_budget)
}

one_play <- function(previous_ledger_entry, max_wager){
  # Create a copy of the input object that will become the output object
  out <- previous_ledger_entry
  out[1, "game_index"] <- previous_ledger_entry[1, "game_index"] + 1
  #game index is increasing with the number of game played. 
  out[1, "starting_budget"] <- previous_ledger_entry[1, "ending_budget"]
  #Each game, player have budget of what they left from last game.
  out[1, "wager"] <- martingale_wager(
    previous_wager = previous_ledger_entry[1, "wager"]
    , previous_outcome = previous_ledger_entry[1, "outcome"]
    , max_wager = max_wager
    , current_budget = out[1, "starting_budget"]
  )
  #To get and record this time's wager in, it call martingale_wager() function with last time's wager, outcomes,max wager and this time's starting budget.
  out[1, "outcome"] <- single_spin()
  # Calling the single_spin() function to get an outcome and record.
  out[1, "ending_budget"] <- out[1, "starting_budget"] + 
    ifelse(out[1, "outcome"] == "red", +1, -1)*out[1, "wager"]
  return(out)
  #the new ending budget will be starting budget plus wager if outcome is red, or minus wager is not red.
}

one_series <- function(
  max_games, starting_budget, winning_threshold, max_wager
){
  # Initialize ledger
  ledger <- data.frame(
      game_index = 0:max_games
    , starting_budget = NA_integer_
    , wager = NA_integer_
    , outcome = NA_character_
    , ending_budget = NA_integer_
  )#this defines the variable types.
  ledger[1, "wager"] <- 1
  ledger[1, "outcome"] <- "red"
  ledger[1, "ending_budget"] <- starting_budget
  # the wager starts with 1.
  for(i in 2:nrow(ledger)){
    #browser()
    ledger[i,] <- one_play(ledger[i-1,], max_wager)
    # i is this time, i-1 refer to the last game. By calling one_play() with last game, this step keeps recording every game played.
    if(stopping_rule(ledger[i,], winning_threshold)) break
  } #this will check if the plyyer reaches stopping rules. If yes then stop.
  # Return non-empty portion of ledger
  ledger[2:i, ]
}

stopping_rule <- function(
  ledger_entry
  , winning_threshold
){
  ending_budget <- ledger_entry[1, "ending_budget"]
  if(ending_budget <= 0) return(TRUE)
  if(ending_budget >= winning_threshold) return(TRUE)
  FALSE
}# The rule to stop are when running out of money or now having money that is equal or more than winning threshold. 

profit <- function(ledger){
  n <- nrow(ledger)
  profit <- ledger[n, "ending_budget"] - ledger[1, "starting_budget"]
  return(profit)
} # The profit equals to last game's budget minus budget at first.

require(magrittr)

# It creates figures of winning and losing with starting budget of $200, winning threshold of $300, maximum 1000 plays and maximum wager of $100
svg(filename = "loser.svg", width=16, height =9)
par(cex.axis=2, cex.lab = 2, mar = c(8,8,2,2), bg = rgb(222, 235, 247, max = 255))
set.seed(1)
ledger <- one_series(1000,200,300,100)
plot(ledger[,c(1,5)], type = "l", lwd = 5, xlab = "Game Index", ylab = "Budget")
dev.off()

svg(filename = "winner.svg", width=16, height =9)
par(cex.axis=2, cex.lab = 2, mar = c(8,8,2,2), bg = rgb(222, 235, 247, max = 255))
set.seed(2)
l2 <- one_series(1000,200,300,100)
plot(l2[,c(1,5)], type = "l", lwd = 5, xlab = "Game Index", ylab = "Budget")
dev.off()
```

## 3 Results

1

![Possible Outcomes: Figure of wining](winner.svg){#id .class width=50% height=50%}

2

![Possible Outcomes: Figure of losing](loser.svg){#id .class width=50% height=50%}
We can see that one can either lose or win with Martingale strategy. Martingale is not 100% profitable.


### Simulation for 1000 times:

This code creates a list to store each round's ending budget and compare with the starting budget $200 to get the earning. There are 1000 rounds and we calculate the average earning by get the means of all earning.
```{r}
# Simulation
walk_out_money <- rep(NA, 1000)
for(j in seq_along(walk_out_money)){
  data = one_series(1000, 200, 300, 100)
  walk_out_money[j] <- (data$ending_budget)[length(data$ending_budget)]
}

# Walk out money distribution
#hist(walk_out_money, breaks = 100)
wom <- data_frame(walk_out_money)
#View(wom)
```

```{r}
# Walk out money distribution
#hist(walk_out_money, breaks = 100)
wom <- data_frame(walk_out_money)
#View(wom)
g1 <- ggplot(wom,aes(x=walk_out_money))+ geom_histogram(bins=100)
g1
```
During 1000 games, The Roulette player either win $300 or lose all money. The player lose nearly half the times.

```{r}
# Estimated probability of walking out with extra cash
mean(walk_out_money > 200)

# Estimated earnings
mean(walk_out_money - 200)
```
The probability that player would win $300 is 51.37% and the average earnining for Roulette is -45.8854. So Martingale Startegy is not profitable and resulting in an average loss.


### Change the winning threshold


```{r}
# Simulation
out <- rep(NA, 5)
win_th <- c(250, 275, 300, 325, 350)
for (i in (1:5)){
  walk_out_money <- rep(NA, 100)
  for(j in seq_along(walk_out_money)){
    data = one_series(1000, 200, win_th[i], 100)
  walk_out_money[j] <- (data$ending_budget)[length(data$ending_budget)]
  }
  out[i] <- mean(walk_out_money - 200)
}

# Walk out money distribution
#hist(walk_out_money, breaks = 100)
```

```{r}
plot(win_th, out)
```

From the above figures, we can see the impact of changing winning threshold from 250 to 350. A lower threshold results in higher walk out money and higher threshold results in a lower walk out budget. So winning threshold has an impact on average earning.

### Average Numbers of Plays

This code creates a list to store how many games are played in each round and there are 1000 rounds and we calculate the average number of play by get the means of all rounds.
```{r}
game_num <- rep(NA, 1000)
for(j in seq_along(game_num)){
  data = one_series(1000, 200, 300, 100)
  game_num[j] <- (data$game_index)[length(data$game_index)]
}

hist(game_num, breaks = 100)
mean(game_num)
```
This is the distribution of the average played game. The average number of plays is 206.884.


### Limitation & Simplication

This simulation is done by computer and does not 100% replicate real-life gambling. People do not follow the stopping rules. For example, people will not stop at winning threshold for they believe in luck. Or casino will not set $100 has maximum wager. These are the uncertain factor of the simulation.


## 4 Conclusions

By changing the input parameter and repeatedly simulate, we find that Martingale betting strategy does not guarantee earnings. A player can reach winning threshold or lose all money. The average earning is negative in this 1000 simulation. So this strategy is not always profitable.  And changing winning threshold shows a higher threshold results in a lower walk out budget. 
