---
title: "Order Statistics"
author: ' Zhuoyi Zhan'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    toc: yes
    number_sections: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
require(tgsify)
library(dplyr)
```


# Introduction

In a sample of size N, Median is middle point of sample.
**median:  g(X1,X2,…,XN)=smallest  x  such that 1N∑i=1NI(Xi≤x)=.5**
The kth order statistic in a statistical sample is the kth-smallest value.


Q: Begin with the median from a sample of N = 200 from the standard normal distribution. Write an R function that is the density function for the median in this sample. Note that the 100th order statistic is approximately the median, and use the order statistic formula discussed in class. Generate a plot of the function.

```{r}
dorder <- function(x, k, n){
  k*choose(n,k)*
    (pnorm(x,0,1))^(k-1)*
    (1-pnorm(x,0,1))^(n-k)*
    dnorm(x,0,1)
}

x<- seq(-5, 15, by = 0.01)
curve(dorder(x, 100, 200), -0.5,0.5)
```

The median is between -0.1 and 0.1.

Q: Write an R function that is the probability function for the median in this sample. Use the order statistic formula discussed in class. Generate a plot of the function.

```{r}
porder <- function(x, k ,n){
  pbinom(k-1, n, pnorm(x, 0, 1), lower.tail = FALSE)
}

x <- seq(-5, 15, by = 0.01)
curve(porder(x,100, 200),-0.5,0.5)


```

Q: Write an R function that is the quantile function for the median in this sample. (You have several options for how to write this function.) Generate a plot of the function.

```{r}
qorder <- function(p){
  out <- p
  for (i in seq_along(p)){
    out[i] <- uniroot(function(x){porder(x,100,200)-p[i]},c(-100,100))$root
  }
  out
}
p<- seq(0.05,0.95,by =0.01)
plot(p,qorder(p),type="l", ylim=c(-2,2))
lines(p, qnorm(p),col="steelblue",lwd=3)
#curve(qorder(x,porder),0.001, 0.9991)
```



Simulate the sampling distribution for the median. Create a plot of the empirical CDF (ECDF). Overlay the plot of the ECDF with a plot of the CDF.

```{r}
samples <- array(rnorm(5000*200, 0, 1),c(5000,200))
medians <- samples %>% apply(1, median)
plot(ecdf(medians))
curve(porder(x,100,200),add=T, lwd=3,col="steelblue")

```

As the simulation number increases, the CDF and ECDF are very similar.

Q: Using the simulated sampling distribution from the previous question, create a histogram (on the density scale). Overlay the histogram with a plot of the density function.

```{r}
samples <- array(rnorm(5000*200, 0, 1),c(5000,200))
medians <- samples %>% apply(1, function(x){sort(x)[100]})
hist(medians, breaks =100, freq=F)
curve(dorder(x,100,200),add=T, lwd=3,col="red")
```

The shape simulated sampling distribution match the density function.

Q: One very common way to compare a random sample to a theoretical candidate distribution is the QQ plot. It is created by ploting quantiles of the theoretical distribution on the x-axis and empirical quantiles from the sample on the y-axis.

```{r}
pts <- ppoints(200)
M <- 5000
N <- 200
samples <- array(rnorm(M*N, 0,1), c(M,N))
medians <- samples %>% apply(1,function(x){sort(x)[100]})
x<- qorder(p)
y<- quantile(medians, prob =p)
plot(x,y,xlim=c(-0.5, 0.5), ylim=c(-0.2,0.2),asp=1)
abline(0,1)
```

The points are onthe y=x line.

Q: Modify the dorder, porder, and qorder functions so that the functions take a new parameter k (for the kt**h order statistic) so that the functions will work for any order statistic and not just the median.

Q: Generate the QQ plot for simulated data from the sampling distribution of the sample max and the theoretical largest order statistic distribution.

Q: Modify the dorder, porder, and qorder functions so that the functions take new parameters dist and ... so that the functions will work for any continuous distribution that has d and p functions defined in R.

Q: Use the newly modified functions to plot the probability and density functions for the sample min (N = 200).