---
title: "Title of Blog Post Deliverable"
author: ' Zhuoyi Zhan'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    toc: yes
    number_sections: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The World Series is the annual championship series of Major League baseball. The winner is decided by a best of 7 match-up. In this blog, rules of probability and discrete probability will be used to answer a series of questions.

## Background

The Braves and the Yankees are competing. Suppose the Braves is the better team with a probability to win of 0.55 and the opponent, Yankees has a probability to win of (1 - 0.55).

In a best of 7 match-up, a team needs to win 4 out 7 to win the World Series.

|     Team     | P(win) | 
|:------------:|:------:|
|  the Braves  |  0.55  | 
|  the Yankees |  0.45  | 

```{r}
library(dplyr)
library(tidyverse)
```
### What is the probability that the Braves win the World Series given that PB=0.55?

Bernouli trial are a series of idependent repeated trail which has two outcomes, "success" and "failure". In this case, failure is lose a match and success is win a match.

A negative binomial distribution models the number of successes in a Bernouli trials before it stops with specific number of failure. dnbinom(x,y,p) is the return the value of probability density that measures probability of x failure before yth success given that probability of success is p. 

To win in World Series, the team need to be the first to win 4 matches. The Braves can win in 4th, 5th, 6th, 7th game because it needs to win 3 out of the first (N-1) games and the last game. 

|Game Index|             Score             |     Probability   |
|:--------:|:-----------------------------:|:-----------------:|
|     4    | win 3 game in 3 and last game | dnbinom(0,4,0.55) | 
|     5    | win 3 game in 4 and last game | dnbinom(1,4,0.55) | 
|     6    | win 3 game in 5 and last game | dnbinom(2,4,0.55) | 
|     7    | win 3 game in 6 and last game | dnbinom(3,4,0.55) |



```{r}
sum(dnbinom(0:3,4,0.55))#the total probability of winning in 4th, 5th, 6th, and 7th game.
```

Given that Pb=0.55, the Brave has a 0.608 probability of winning the World Series. The probability is higher than winning a single match.

### Braves win the World Series given that Pb is x

Now we assume P(brave win in a game) is between 0.5 to 1.

First I generate a series of Pb with gaps of 0.01, and a vector to store probability of winning World Series. Then I use loop to calculate probability with given Pb and store in vector.

pnbinom(), cumulative possibilty function, is equivalent to sum(dnbinom())

```{r}
prob_th <- c(seq(0.5,1,by =0.01))
out <- rep(NA, length(prob_th))
for (i in (1:length(prob_th))){
  out[i] <- pnbinom(3,4,prob_th[i])
}
plot(prob_th,out,type="l",xlab="P(the Braves Win A Game)", ylab ="P(the Braves Win World Series)", main = "Probabilty of the Braves Win World Series")
```

This is a figure of Pb vs P(the Braves Win World Series). There is a positive non-linear relationship between pb and the Braves win the World Series. When the probabilityof winnig a single match increases, the probability that the braves win the World Series increases.

### In best-of-X series, what is the shortest series length so that P(Braves win World Series|PB=.55)≥0.8

Now we can change the total game to be any number. Out of N games, the team needs to win (N+1)/2 games. So I add a loop on game number from 1 to 999. It keeps track of the game number until P(Braves win World Series|PB=.55) is greater than 0.8.

```{r}
b_win <-0.55
game_seq <- seq(1, 999, 2)
for (i in 1:length(game_seq)){
  p_win <- pnbinom((game_seq[i]-1)/2,(game_seq[i]+1)/2, b_win)
  if (p_win>=0.8){
    shortest <- game_seq[i]
    p_shortest <- p_win
    break
  }
}
shortest
p_shortest
  

```

When Pb = 0.55, the shortest series length is 71 when P(Braves win World Series|PB=.55) is 0.802. In reality, playing 71 games is impossible and  a 80% probability in winning the world series is unreachable.

### What is the shortest series length so that P(Braves win World Series|PB = x) ≥ 0.8? This will be a figure (see below) with PB on the x-axis and series length is the y-axis.

In this code, a sequence for Pb from 0.5 to 1 and a sequence for games from 1 to 9999 are used. In this way, we can calculate P(brave win the World) with changing Pb and game number with loops.

```{r}
prob_seq <- seq(0.51,1,0.01)
out <- rep(NA, length(prob_seq))
game_seq <- seq(1, 9999, 2)
for (k in 1:length(prob_seq)){
  for (i in 1:length(game_seq)){
    p_win <- pnbinom((game_seq[i]-1)/2,(game_seq[i]+1)/2, prob_seq[k])
    if (p_win>=0.8){
     shortest <- game_seq[i]
     break
    }
  }
  out[k] <- shortest
}

plot(prob_seq, out, type="l",xlab="P(the Braves Win A Game)", ylab ="Shortest Series Length", main = "Shortest Series when P(Braves win Series|PB = x) ≥ 0.8")
```

There is a negative relationship between Pb and shortest series length. As Pb increase, the shortest series length decreases. When the probability of winning a game is high, the less game needs to be played

### Calculate P(PB = 0.55|Braves win World Series in 7 games) under the assumption that either PB = 0.55 or PB = 0.45. Explain your solution.

| Situation | Probability | 
|:---------:|:-----------:|
|  Pb=0.55  |     0.5     | 
|  Pb=0.45  |     0.5     | 

Bayes rule should be used here. 
$$P(A|B) = P(B|A)*P(A)/P(B)$$
$$ P(Pb=0.55|Braves Win World Series) = P(Braves win world serie|Pb=0.55)*P(Pb=0.55)/(P(Braves win world games|Pb =0.55)*0.5 +(P(Braves win world games|Pb =0.45)*0.5 $$

```{r}
(dnbinom(3,4,0.55)*0.5) /(dnbinom(3,4,0.55)*0.5 + dnbinom(3,4,0.45)*0.5)
```

Given the Braves win world series, the probability of winning a single match is 0.55 which is the same with probability of winning a single game without conditions.
